Transparent Changes

### 2022-04-28 thesis/discussion.Rmd
Continue writing on thesis

>>>>

<!-- if report - diverse strategies -->
<!-- most (multiple) binned time averages: necessary for mixing chamber, but bbb has more information! -->
<!-- current practice is in contrast to recommendations (Robergs)! -->

<!-- variety of calculation intervals - different results (example)-->
<!-- comparing results for strategies -->

<!-- recommendation for choosing a strategy: ... -->

<!-- guidelines for reporting: ... -->

### 2022-04-28 thesis/methods.Rmd
Continue writing on thesis

The work presented in this thesis was preregistered before the start of the project on [OSF](). Any deviations from the preregistration are indicated in the 'Transparent Changes' document (LINK/Appendix). All data and code can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).

@@

<!-- Detailed Aim of review -->

@@

To assess the current state of data processing methods for determining V̇O~2max~ in exercise science I systematically screened recent literature. The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science](). The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword. Table 1 shows the exact search terms.

@@

knitr::kable(out, 
  col.names = c("Source", "Search String"), 
  caption = "Search strings for the systematic scoping review.")
```

The search results from both data bases were joined at checked for the presence of a DOI. Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step. After the removal of duplicates I conducted an automated title scanning to remove results that are likely no original research articles. Table 2 displays the exclusion terms during the automated title screening.

@@

The abstracts from the articles included in the random sample were blinded for scanning. This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the articles (see Figure 2). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3. When the screeners disagreed in their assessments, conflicts were resolved by discussion.

@@

-   type of metabolic cart/data basis (breath-by-breath, mixed chamber, ...)

@@

All extracted data is available in the Appendix A and under XXXLINK.

@@

<!-- presentation of data -->

@@

A total of N = 72 exercise test were analysed for this study. The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times). The V̇O~2max~ tests were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol. Participants run on a treadmill (XXX) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up. After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently velocity increased by 0.15 m/s every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

@@

The spiro Package for R [@nolte2022a] processed the raw gas exchange data. It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data. Table 4 displays a brief summary exemplary function parameters used for the analysis. The full analysis script is available at XXXLINK.

>>>>

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](). Any deviations from the preregistration are indicated in the 'Transparent Changes' document (LINK/Appendix). Major deviations will also explicitly name be named within the methods section. All data and code of this research project can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).

@@

The aim of the Scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since the V̇O~2max~ determination is too common to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies was extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.

The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A contains the corresponding reporting checklist.

@@

The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science](). The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keyword. Table 1 shows the exact search terms used.

@@

knitr::kable(out, 
  col.names = c("Source", "Search String"), 
  caption = "Search strings for the systematic scoping review."
)
```

The search results from both data bases were joined and checked for the presence of a DOI. Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step. After the removal of duplicates I conducted an automated title scanning to remove results that were likely no original research articles. Table 2 displays the exclusion terms during the automated title screening.

@@

The abstracts from the articles included in the random sample were blinded for scanning. This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix XXX for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3. When the screeners disagreed in their assessments, conflicts were resolved by discussion.

@@

-   type of metabolic cart/data basis (breath-by-breath, mixing chamber, ...)

@@

All extracted data is available in the Appendix XXX and under XXXLINK.

@@

The extracted data is presented in a purely descriptive way. I calculated relative and absolute proportions of the different extracted variables.

@@

A total of N = 72 exercise test were analysed for this study. The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times). The V̇O~2max~ tests were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol. Participants run on a treadmill (XXX) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up. After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

@@

The spiro Package Version XXX for R [@nolte2022a] processed the raw gas exchange data. It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data. Table 4 displays a brief summary of exemplary functions and parameters used for the analysis. The full analysis script is available at XXXLINK.

```{r spiro-function, echo = FALSE}
spiro_functions <- data.frame(
  proc_type = c("Moving Time Average", "Binned Time Average", "Multiple Binned Time Average", "Moving Breath Average", "Digital Filtering"),
  func = c("`spiro_max()`", "`spiro_summary()`", "`spiro_summary(); filter()`", "`spiro_max()`", "`spiro_max()`"),
  parameter = c("`smooth = 30`", "XXX", "XXX", '`smooth = "10b"`', '`smooth = "XXX"')
)

knitr::kable(spiro_functions)
```

### 2022-04-28 thesis/results.Rmd
Continue writing on thesis

@@

![A plot](../preregistration/example_plot.png)

>>>>

Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in Figure 1). Out of the random sample (n = 500), 244 articles were included in the final analysis.

![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png)

Reporting practices of the methodology of gas exchange measures differed widely within the literature (see Table 4). More than half (51.8%) of the articles did not report any information on their data processing strategy. Only one in twenty articles (5.4%) provided a rationale for their used strategy.

```{r reporting-table, echo=FALSE}
load("../data/reporting.Rda")
t <- purrr::map_dfr(reporting, scales::percent, accuracy = 0.1)

t[["pre"]] <- paste0(t[["pre"]], "*")
t[["software"]] <- paste0(t[["software"]], "*")

knitr::kable(
  t,
  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
  caption = "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing. *only examined within the subgroup of studies using breath-by-breath measurements",
  align = "r"
)
```

Out of the authors that provided information and collected breath-by-breath measurements most (79.5%) utilized binned averages to determine V̇O~2max~. Moving time averages, or breath-based averages were uncommon (see Figure 2). No study used digital filtering methods to determine V̇O~2max~.

![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png)

The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see Figure 3). 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20s) and longer (60s) periods.

![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png)

@@

![VO2max varies by data processing strategy. Values are expressed relative to the VO2max from a 30-second binned average --- the most common strategy as determined by the review. Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile. Using moving average leads to systematically higher VO2max values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to changes in VO2max as large as 5%.](../plots/comparison.png)

![Respiratory Rates peak around 60min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72). .](../plots/rr.png)
